[![Travis-CI Build Status](https://travis-ci.org/arendsee/rhmmer.svg?branch=master)](https://travis-ci.org/arendsee/rhmmer)
[![Coverage Status](https://img.shields.io/codecov/c/github/arendsee/rhmmer/master.svg)](https://codecov.io/github/arendsee/rhmmer?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/rhmmer)](https://cran.r-project.org/package=rhmmer)
[![CRAN downloads](http://cranlogs.r-pkg.org/badges/rhmmer)](http://cran.rstudio.com/web/packages/rhmmer/index.html)
[![total downloads](http://cranlogs.r-pkg.org/badges/grand-total/rhmmer)](http://cranlogs.r-pkg.org/badges/grand-total/rhmmer)
[![DOI](https://zenodo.org/badge/114339623.svg)](https://zenodo.org/badge/latestdoi/114339623)

# rhmmer

[HMMER](hmmer.org) is a powerful package for profile HMM analysis. If you want
to interface with the web server through R, for example to search for domains
in a small number of proteins, consider using the Bio3D package.`rhmmer` is
specifically designed for working with the standalone HMMER tool.

## Installation

`rhmmer` is available on CRAN

```{r, eval=FALSE}
install.packages('rhmmer')
```

Alternatively, you may install the github development version

```{r, eval=FALSE}
library(devtools)
install_github('arendsee/rhmmer')
```

## Examples

`rhmmer` currently exports exactly two functions: `read_domtblout` and
`read_tblout`. These read the `hmmscan` outputs specified by the `--domtblout`
and `--tblout` arguments, respectively.

The `domtblout` files have the format:

```
#                                                                            --- full sequence --- -------------- this domain -------------   hmm coord   ali coord   env coord
# target name        accession   tlen query name           accession   qlen   E-value  score  bias   #  of  c-Evalue  i-Evalue  score  bias  from    to  from    to  from    to  acc description of target
#------------------- ---------- ----- -------------------- ---------- ----- --------- ------ ----- --- --- --------- --------- ------ ----- ----- ----- ----- ----- ----- ----- ---- ---------------------
Rer1                 PF03248.12   171 AT2G18240.1          -            221   2.3e-61  206.4  10.4   1   2   4.3e-65   7.1e-61  204.8  11.2     4   168    24   181    21   183 0.96 Rer1 family
Rer1                 PF03248.12   171 AT2G18240.1          -            221   2.3e-61  206.4  10.4   2   2      0.43   7.2e+03   -3.6   0.0    55    67   187   199   184   217 0.69 Rer1 family
DUF4220              PF13968.5    352 AT5G45530.1          -            798   4.6e-78  263.0   0.0   1   1   1.5e-81   1.3e-77  261.5   0.0     1   344    51   396    51   427 0.78 Domain of unknown function (DUF4220)
DUF594               PF04578.12    55 AT5G45530.1          -            798   4.7e-24   83.6   1.5   1   1   1.3e-27   1.1e-23   82.4   1.5     3    55   726   778   724   778 0.96 Protein of unknown function, DUF594
DEAD                 PF00270.28   176 AT1G27880.2          -            890   6.7e-20   71.5   0.1   1   1   3.4e-23   1.9e-19   70.0   0.1     2   171   272   433   271   438 0.83 DEAD/DEAH box helicase
...
#
# Program:         hmmscan
# Version:         3.1b2 (February 2015)
# Pipeline mode:   SCAN
# Query file:      five.faa
# Target file:     /home/z/db/Pfam-A.hmm
# Option settings: hmmscan --tblout x.tblout --domtblout x.domtblout --pfamtblout x.pfamtblout --noali /home/z/db/Pfam-A.hmm five.faa 
# Current dir:     /home/z/src/git/rhmmer/tests/testthat/sample-data
# Date:            Fri Dec 15 02:09:00 2017
# [ok]
```

This is tricky to parse. It is mostly space delimited, but spaces appear freely
in the `description of target` column. The column names, as given, cannot be
directly used since they 1) contain illegal characters and 2) are not unique
unless information from two rows is considered (e.g. `ali_from` versus
`env_from`). The metadata at the end of the file I do not currently extract,
though I will likely add handling for this in the future.

```{r}
library(rhmmer)
domtblout <- system.file('extdata', 'example.domtblout.txt', package='rhmmer')
read_domtblout(domtblout)
```

The `tblout` output is fairly similar and presents the same parsing difficulties:

```
#                                                               --- full sequence ---- --- best 1 domain ---- --- domain number estimation ----
# target name        accession  query name           accession    E-value  score  bias   E-value  score  bias   exp reg clu  ov env dom rep inc description of target
#------------------- ---------- -------------------- ---------- --------- ------ ----- --------- ------ -----   --- --- --- --- --- --- --- --- ---------------------
Rer1                 PF03248.12 AT2G18240.1          -            2.3e-61  206.4  10.4   7.1e-61  204.8  11.2   1.3   2   0   0   2   2   2   1 Rer1 family
DUF4220              PF13968.5  AT5G45530.1          -            4.6e-78  263.0   0.0   1.3e-77  261.5   0.0   1.7   1   1   0   1   1   1   1 Domain of unknown function (DUF4220)
DUF594               PF04578.12 AT5G45530.1          -            4.7e-24   83.6   1.5   1.1e-23   82.4   1.5   1.7   1   0   0   1   1   1   1 Protein of unknown function, DUF594
DEAD                 PF00270.28 AT1G27880.2          -            6.7e-20   71.5   0.1   1.9e-19   70.0   0.1   1.8   1   0   0   1   1   1   1 DEAD/DEAH box helicase
Helicase_C           PF00271.30 AT1G27880.2          -            3.2e-18   66.0   0.0     3e-17   62.9   0.0   2.5   2   0   0   2   2   2   1 Helicase conserved C-terminal domain
ResIII               PF04851.14 AT1G27880.2          -             0.0012   18.8   0.1    0.0084   16.0   0.0   2.4   2   1   0   2   2   2   1 Type III restriction enzyme, res subunit
...
#
# Program:         hmmscan
# Version:         3.1b2 (February 2015)
# Pipeline mode:   SCAN
# Query file:      five.faa
# Target file:     /home/z/db/Pfam-A.hmm
# Option settings: hmmscan --tblout x.tblout --domtblout x.domtblout --pfamtblout x.pfamtblout --noali /home/z/db/Pfam-A.hmm five.faa 
# Current dir:     /home/z/src/git/rhmmer/tests/testthat/sample-data
# Date:            Fri Dec 15 02:09:00 2017
# [ok]
```

```{r}
tblout <- system.file('extdata', 'example.tblout.txt', package='rhmmer')
read_tblout(tblout)
```


## What `rhmmer` does and doesn't

What `rhmmer` currently does

 * parse HMMER output (tblout and domtblout) into tidy data frames

What `rhmmer` may do in the future

 * provide visualization or analysis of HMMER results 

 * provide simple wrappers for calling some of the HMMER functions

 * parse metadata from the comments in HMMER output files

 * read pfamtblout and the raw output (with alignments)

 * read the HMM models

 * download databases for you (e.g. PFAM)

 * access precompiled PFAM results

What `rhmmer` will never do

 * interface with the HMMER web server (use Bio3D)

 * duplicate the whole HMMER api (just use the CLI tool)


## TODO

 [ ] Rewrite the parser in C++

 [ ] Parse out the file metadata
