---
title: "Protein domain identification"
author: "Zebulun Arendsee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Protein domain identification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

One of the main uses of HMMER is identifying protein domains. You can search
individual proteins easily on the HMMER website (http://hmmer.org/). But for
higher throughput analysis, a domain database, such as PFAM, needs to be
downloaded. The PFAM database is freely available from EBI. You can retrieve
the current PFAM release from here: 

```
ftp://ftp.ebi.ac.uk/pub/databases/Pfam/current_release/Pfam-A.hmm.gz
```

This file can then be unzipped and pressed with `hmmpress` (a tool in the HMMER
suite). Here is a shell script for the process (assuming a UNIX environment):

```{sh, eval=FALSE}
wget ftp://ftp.ebi.ac.uk/pub/databases/Pfam/current_release/Pfam-A.hmm.gz
gunzip Pfam-A.hmm.gz
hmmpress Pfam-A.hmm
```

It should only take a few minutes to install and build the database. On
success, the following files are generated: `Pfam-A.hmm`, `Pfam-A.hmm.h3f`,
`Pfam-A.hmm.h3i`, `Pfam-A.hmm.h3m`, and `Pfam-A.hmm.h3p`.

In this vignette, we we will explore the domains of the rhino mitochondrial
proteins. These sequences are included in the package data.


```{r}
prot <- system.file('extdata', 'rhino.faa', package='rhmmer')
```

Now we can search these proteins against the PFAM database. You could run the
HMMER program from within R through `system2`, but here I will just give the
shell command.


```{sh, eval=FALSE}
hmmscan --tblout rhino.tblout --domtblout rhino.domtblout Pfam-A.hmm rhino.faa
```

This spills a bunch of text to STDOUT and creates two new files:
`rhino.domtblout` and `rhino.tblout`. In case you are not executing all the
installation and database building steps, I've included these outputs in the
package data:

```{r}
domtblout_file <- system.file('extdata', 'rhino.domtblout', package='rhmmer')
tblout_file <- system.file('extdata', 'rhino.tblout', package='rhmmer')
```

Now we get to the part `rhmmer` plays, which is to read these files into a tidy
format: 


```{r}
library(rhmmer)
domtblout <- read_domtblout(domtblout_file)
tblout <- read_tblout(tblout_file)
```

For a description of all the columns, it is best to go to the [HMMER
manual](http://eddylab.org/software/hmmer3/3.1b2/Userguide.pdf).

Here are a few quick summaries of the results

```{r, results="asis"}
library(dplyr)
library(magrittr)
library(knitr)
domtblout %>%
  filter(domain_ievalue < 1e-6) %>%
  select(domain_name, domain_accession, description) %>%
  unique %>%
  arrange(domain_name) %>%
  kable(caption="Summary of significant domains")
```

```{r, results="asis"}
tblout %>%
  filter(best_domain_evalue < 0.001) %>%
  select(query_name, domain_name) %>%
  kable(caption="Map of query protein to domain")
```
